\chapter{Datasets}

We test our hypotheses on two datasets: MUSIN-G\cite{musing} and what we call BCMI-training\cite{bcmi}.

\section{MUSIN-G}

Is a dataset of joint music and EEG recordings of 20 non-musician subjects, aged 22-28 years.
Each one was presented with every one of 12 songs of various genres from Indian Folk to Electronic Dance.

\section{BCMI-training}

BCMI-training is a dataset of joint music and EEG recordings of 10 participants aged 19-30.
It is part of a series of datasets collected during a series of studies that investigate human reaction to affective musical stimuli and whether reported emotionality can be influenced by music\cite{nicolaou}.
The music played in this study is fully computer generated and no track repeats across the dataset.
The musical snippets are tagged with emotionality labels (9 possibilities on valence/arousal grid) which are inputs to the music generation algorithm.
In the study, subjects listened to a musical snippet of about 20 seconds in length and immediately after to another snippet, but of
different emotionality.
For our purpose we split such trials into two separate recordings, one for each snippet.

\section{Dataset processing}

Adding to the methods, we describe how a dataset is loaded to a study using our library.

We unify various analyzed datasets under the `EEGMusicDataset` class, which provides a common interface to access the data and metadata of the recordings.
Then, various preprocessing functions are applied to the data, the evaluation of which is lazy.
Lastly, `StratifiedSamplingDataset` wrapper gets applied which samples the data in a stratified way, as described in the methods chapter.

\begin{verbatim}
ds = EEGMusicDataset.load_ondisk(Path("..."))

ds = MappedDataset(ds, lambda x: 
  ica_band_power_trial(
    rereference_trial(
      filter_trial(
        notch_filter_trial(x, 50.0), l_freq=0.5, h_freq=50.0
      )
    ),
    n_components=12,
    bands = [(0.5, 4), (4, 8), (8, 13), (13, 30), (30, 45)],
    window_sec= 2.0,
    hop_sec = 0.1,
  )
)

ds = StratifiedSamplingDataset(ds, n_strata=10, trial_length_secs=Fraction(3, 1))

\end{verbatim}

% musing


% grupa bcmi:
%  - opis eksperymentu
%  - używamy treningowego, choć można połączyć